High performance computing for global optimization problems

### 2018-05-25 2018/ITNT/ITNT2018.tex
Semi-final

\address{National Research Lobachevsky State University of Nizhny Novgorod, Gagarin avenue 23, 
Nizhny Novgorod, Russia, 603950}

@@

\begin{abstract}
In the present work, the multiextremal optimization problems and a high-
performance parallel algorithm for solving these ones are considered. The 
investigation of the algorithm scalability has been carried out on the 
problem class, in which the computation costs of the functions depended on 
the iteration point. The algorithm proposed in the present work can utilize 
the CPUs (for solving more complex subproblems) as well as the GPUs (for 
solving the simple subproblems). The results of numerical experiments 
demonstrating the speedup when solving a series of multiextremal constrained 
problems are presented.
\end{abstract}

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
optimized function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate a nonuniform grid in the search domain, denser near the 
global minimizer and more sparse far away from this one. In the 
present paper, the results of scalability investigation of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system \cite{Globalizer,Globalizer1} are presented.

@@

An assumption on different costs of computing the problem functions subject to 
various components of the vector of parameters is a novel element investigated in the present work. 
It is assumed that it is possible to select the ``difficult'' and ``easy'' parts 
of the functions connected with each other. An example of a problem of this type has been considered in \cite{Strongin2017}.
Then, the solving of the ``difficult'' part of the problem (for which the execution of time-consuming algorithms is required to conduct the trials) can be performed on CPU using an efficient index global optimization algorithm. The ``easy'' part of the problem (the execution of which doesn't require the complex algorithms 
and can be transferred to an accelerator easily) can be solved on GPU using 
the uniform grid technique. Obviously, in this case a major portion of the trials 
will be carried out on GPU, and the minor one will be carried out on CPU. However, 
because of the difference in the difficulties of the subproblems solved on 
different units, a speedup of the algorithm as a whole can be expected. 
The proposed approach has been implemented in Globalizer parallel software 
system for solving the global optimization problem developed in Lobachevsky 
State University of Nizhny Novgorod.

@@

Let us consider the $N$-dimensional global optimization problem
\begin{equation}\label{problem}
\varphi(y^\ast)=\min{\left\{\varphi(y): y \in D, \; g_i(y)\leq 0, \; 1 \leq i 
\leq m\right\}}
\end{equation}
with search domain
\[
D=\left\{y\in R^N: a_j\leq y_j \leq b_j, 1\leq j \leq N \right\}.
%D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}.
\]

@@

Thus, a \textit{trial} at a point $y^k \in D$ executed at the $k$-th
iteration of the algorithm will consist in computing the values 
$g_1(y),...,g_\nu(y),$ where the
index $\nu \leq m$ is determined by the conditions
\[
	g_i(y^k )\leq 0, \; 1 \leq i < \nu, \; g_\nu(y^k)>0, \; \nu \leq m.
\]
The occurrence of the first violation of the constraint terminates the
trial. In the case when the point $y^k$ is a feasible one, i.e. when
$y \in Q_{m+1}$, the trial includes the computation of the values of
all the functions of the problems and the index is assumed to be
$\nu=m+1$. The pair of values
\[
\nu=\nu(y^k), \; z^k=g_\nu(y^k)
\]
is a \textit{trial result}.

@@

Rule 3. Compute the current lower estimates,
\begin{equation}\label{Rule_Mu}
\mu = \max\left\{ \frac{\left|z_i-z_j\right|}{ x_i - x_j }, \; i,j \in I_\nu, \; i>j \right\},
\end{equation}
for the unknown Lipschitz constants $L_\nu$ of the functions $g_\nu(x),1
\leq \nu \leq m+1$. If a set $I_\nu$ contains less than two elements, or
if $\mu_\nu$ is equal to zero, then assume $\mu_\nu=1$.

@@

Nested optimization scheme is based on relation (see \cite{Strongin2013})
\[
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}...\min_{u_M\in D_M 
}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq m\right\}},
\]
which allows replacing the solving of multidimensional problem 
(\ref{problem}) by solving a family of subproblems related to each other 
recursively.
Here we consider vector $y$ as a vector of block variables
\[
y=(y_1,y_2,...,y_N)=(u_1,u_2,...,u_M),
\]
where the $i$-th block variable $u_i$ is a vector of $N_i$ components of 
vector $y$, taken serially, i.e. $u_1=(y_1,y_2,...,y_{N_1})$, 
$u_2=(y_{N_1+1},y_{N_1+2},...,y_{N_1+N_2})$,..., $u_M=(y_{N-N_M+1},y_{N-
N_M+2},...,y_{N})$, at that $N_1+N_2+...+N_M=N$. 
The subdomains $D_i, 1 \leq i \leq M$, are projections of initial search 
domain $D$ onto the subspaces corresponding to the variables $u_i, 1 \leq i 
\leq M$.

The number of vectors $M$ and the quantity of components in each vector $N_1, N_2,...,N_M$ are the parameters of the nested optimization scheme and can be used for the forming of the subproblems with necessary properties. For example, if  $M=N$ i.e. $u_i=y_i, 1\leq i \leq N,$ each nested subproblem is a one-dimensional one. If $M=1$, i.e. $u=u_1=y$ , the solving of the problem is equivalent to solving this one using a single evolvent mapping $[0,1]$ into $D$; the nested subproblems are absent.

@@

The organization of parallel computing with the use of the recursive 
optimization scheme has been considered in details for the shared/distributed 
memory as well as for the accelerators in \cite{Sysoyev,BarkalovLebedev}. However, in these works the 
problems were considered, in which the time of the trial execution didn't 
depend on the trial point. Here, the problem is considered assuming that in 
the function $\varphi(y_1,...,y_N)$ it is possible to select more difficult 
part $f(y_1,...,y_s)$ (depending on a part of the parameters only) and a simpler 
part $g(y_1,...,y_N)$ (depending on all problem parameters), for 
example, $\varphi(y)= f(y)g(y)$.

The difficult part of the function implies performing some  
computations related to the numerical simulation, which can be performed on the 
CPUs only. The simple part doesn't imply the complex computations and can be 
computed on an accelerator, for example, GPU. For solving a problem with such 
a structure, one can apply the parallel recursive optimization scheme with 
the use CPU at the upper nesting level and GPU at the lower one.

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Globalizer,Globalizer1}. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{BarkalovGergelLebedev,BarkalovGergel} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

@@

Let us conduct the next experiment employing $p$ nodes each including three 
graphic accelerators. Thus, at $p=32$ total 96 GPU accelerators were employed 
each including 2688 CUDA cores. In tables \ref{tab:2} and \ref{tab:3} the average time and the speedup with respect to the run 
on a single node are presented.

@@

Computational experiments were carried out on a high-performance cluster of 
Lobachevsky State University of Nizhny Novgorod. The cluster node included 
two Intel Sandy Bridge E5-2660 2.2 GHz CPUs, 64 Gb RAM, and three NVIDIA 
Kepler K20 GPUs (2688 CUDA cores, 6 Gb GDDR5). 
The results of experiments demonstrate that the parallel index algorithm combined with the nested optimization scheme provides a good speedup on the considered problem class.

@@

\bibitem{Sysoyev}
Sysoyev A V, Barkalov K A, Gergel V P and Lebedev I G 2015 \textit{CEUR Workshop Proceedings} \textbf{1482} 61--68

@@

\bibitem{Gergel2005}
Gergel V P and Strongin R G 2005 \textit{Lect. Notes Comput. Science} \textit{2763} 76--88

\bibitem{BarkalovGergelLebedev}
Barkalov K, Gergel V, Lebedev I 2015 \textit{Lect. Notes Comput. Science} \textbf{9251} 307--318

\bibitem{BarkalovGergel}
Barkalov K and Gergel V 2016 \textit{J. Glob. Optim.} \textbf{66(1)} 3--20

>>>>

\address{Lobachevsky State University of Nizhni Novgorod, Gagarin avenue 23, 
Nizhnш Novgorod, Russia, 603950}

@@

\begin{abstract}
In the present work, the multiextremal optimization problems and a high-
performance parallel algorithm for solving these ones are considered. The 
investigation of the algorithm scalability has been carried out on the 
problem class, in which the computation costs of the function value evaluations can be varied at different iteration points. The algorithm proposed in the present work can utilize 
the CPUs (for solving more complex subproblems) as well as the GPUs (for 
solving the simple subproblems). The results of numerical experiments 
demonstrating the speedup when solving a series of multiextremal constrained 
problems are presented.
\end{abstract}

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
 function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate a nonuniform grid in the search domain, denser near the 
global minimizer and more sparse far away from this one. In the 
present paper, the results of scalability investigation of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhni Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system \cite{Globalizer,Globalizer1} are presented.

@@

An assumption on different costs of computing the problem functions subject to 
various components of the vector of parameters is a novel element investigated in the present work. 
It is assumed that it is possible to select the ``difficult'' and ``easy'' parts 
of the functions connected with each other. An example of a problem of this type has been considered in \cite{Strongin2017}.
Then, the solving of the ``difficult'' part of the problem (for which the execution of time-consuming algorithms is required to conduct the trials) can be performed on CPU using an efficient index global optimization algorithm. The ``easy'' part of the problem (the execution of which doesn't require the complex algorithms 
and can be transferred to an accelerator easily) can be solved on GPU using 
the uniform grid technique. Obviously, in this case a major portion of the trials 
will be carried out on GPU, and the minor one will be carried out on CPU. However, 
because of the difference in the difficulties of the subproblems solved on 
different units, a speedup of the algorithm as a whole can be expected. 
The proposed approach has been implemented in Globalizer parallel software 
system for solving the global optimization problem developed in Lobachevsky 
State University of Nizhni Novgorod.

@@

Let us consider the $N$-dimensional global optimization problem
\begin{equation}\label{problem}
\varphi(y^\ast)=\min{\left\{\varphi(y): y \in D, \; g_i(y)\leq 0, \; 1 \leq i 
\leq m\right\}}
\end{equation}
with the search domain
\[
D=\left\{y\in R^N: a_j\leq y_j \leq b_j, 1\leq j \leq N \right\}.
%D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}.
\]

@@

Thus, a \textit{trial} at a point $y^k \in D$ executed at the $k$-th
iteration of the algorithm will consist in computing the values 
$g_1(y^k),...,g_\nu(y^k),$ where the
index $\nu \leq m$ is determined by the conditions
\[
	g_i(y^k )\leq 0, \; 1 \leq i < \nu, \; g_\nu(y^k)>0, \; \nu \leq m.
\]
The occurrence of the first violation of the constraint terminates the
trial. In the case when the point $y^k$ is a feasible one, i.e. when
$y \in Q_{m+1}$, the trial includes the computation of the values of
all the functions of the problems and the index is assumed to be
$\nu=m+1$. The pair of values
\[
\nu=\nu(y^k), \; z^k=g_\nu(y^k)
\]
is a \textit{trial result}.

@@

Rule 3. Compute the current lower estimates
\begin{equation}\label{Rule_Mu}
\mu_\nu = \max\left\{ \frac{\left|z_i-z_j\right|}{ x_i - x_j }, \; i,j \in I_\nu, \; i>j \right\}, \; 1 \leq \nu \leq m+1,
\end{equation}
for the unknown Lipschitz constants $L_\nu$ of the functions $g_\nu(x),1
\leq \nu \leq m+1$. If a set $I_\nu$ contains less than two elements, or
if $\mu_\nu$ is equal to zero, then assume $\mu_\nu=1$.

@@

Nested optimization scheme is based on relation (see \cite{Strongin2013})
\[
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}...\min_{u_M\in D_M 
}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq m\right\}},
\]
which allows replacing the solving of multidimensional problem 
(\ref{problem}) by solving a family of subproblems related to each other 
recursively.
Here we consider vector $y$ as a vector of block variables
\[
y=(y_1,y_2,...,y_N)=(u_1,u_2,...,u_M),
\]
where the $i$-th block variable $u_i$ is a vector of $N_i$ components of 
vector $y$, taken serially, i.e. $u_1=(y_1,y_2,...,y_{N_1})$, 
$u_2=(y_{N_1+1},y_{N_1+2},...,y_{N_1+N_2})$,..., $u_M=(y_{N-N_M+1},y_{N-
N_M+2},...,y_{N})$, and $N_1+N_2+...+N_M=N$. 
The subdomains $D_i, 1 \leq i \leq M$, are projections of initial search 
domain $D$ onto the subspaces corresponding to the variables $u_i, 1 \leq i 
\leq M$.

The number of vectors $M$ and the quantity of components in each vector $N_1, N_2,...,N_M$ are the parameters of the nested optimization scheme and can be used to form the subproblems with necessary properties. For example, if  $M=N$ i.e. $u_i=y_i, 1\leq i \leq N,$ each nested subproblem is a one-dimensional one. If $M=1$, i.e. $u=u_1=y$ , the solving of the problem is equivalent to solving this one using a single evolvent mapping $[0,1]$ into $D$; the nested subproblems are absent.

@@

The organization of parallel computing with the use of the recursive 
optimization scheme has been considered in details for the shared/distributed 
memory as well as for the accelerators in \cite{BarkalovGergelLebedev2016,BarkalovLebedev}. However, in these works the 
problems were considered, in which the time of the trial execution didn't 
depend on the trial point. Here, the problem is considered assuming that in 
the function $\varphi(y_1,...,y_N)$ it is possible to select more difficult 
part $f(y_1,...,y_s)$ (depending on a part of the parameters only) and a simpler 
part $g(y_1,...,y_N)$ (depending on all problem parameters), for 
example, $\varphi(y_1,...,y_N)= f(y_1,...,y_s)g(y_1,...,y_N)$.

The difficult part of the function implies performing some  
computations related to the numerical simulation, which can be performed on the 
CPUs only. The simple part doesn't imply the complex computations and can be 
computed on an accelerator, for example, GPU. For solving a problem with such 
a structure, one can apply the recursive optimization scheme in parallel with 
the use CPU at the upper nesting level and GPU at the lower one.

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhni Novgorod \cite{Globalizer,Globalizer1}. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{GergelLebedev,GergelSidorov} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

@@

Let us conduct the next experiment employing $p$ nodes each including three 
graphic accelerators. Thus, at $p=32$ total 96 GPU accelerators were employed 
each including 2688 CUDA cores; 258048 CUDA cores in all. In tables \ref{tab:2} and \ref{tab:3} the average time and the speedup with respect to the run 
on a single node are presented.

@@

The results of experiments demonstrate that the parallel index algorithm combined with the nested optimization scheme provides a good speedup on the considered problem class.

@@

\bibitem{BarkalovGergelLebedev2016}
Barkalov K, Gergel V and Lebedev I  2016 \textit{AIP Conf. Proc.} \textbf{1738} 400006

@@

\bibitem{Gergel2005}
Gergel V P and Strongin R G 2003 \textit{Lect. Notes Comput. Science} \textit{2763} 76--88

\bibitem{GergelLebedev}
Gergel V and Lebedev I 2015 \textit{Proc. Comput. Science} \textbf{66} 53--62

\bibitem{GergelSidorov}
Gergel V and Sidorov S 2015 \textit{Lect. Notes Comput. Science} \textbf{9251} 505--515
